---
title: "Hate Crimes in Toronto"
author: "Shamayla Durrin Islam"
date: "22 September 2024"
format: pdf
bibliography: references.bib
link-citations: true
engine: knitr
---

## 1 Introduction

```{r}
library(here)
library(dplyr)
library(grid)

```

```{r}
cleaned_data <- read.csv(here("data/analysis_data/cleaned_data.csv"))
```

## 2 Data

## 2.1 Raw Data

The dataset used in this analysis includes all verified hate crime occurrences investigated by the Hate Crime Unit by reported date since 2018. The Hate Crime categories (bias categories) include Age, Mental or Physical Disability, Race, Ethnicity, Language, Religion, Sexual Orientation, Gender, and Other Similar Factors. The dataset is provided at the offence and/or occurrence level, meaning that one occurrence may have multiple bias categories associated with it. The data does not include occurrences deemed unfounded or classified as hate incidents.

The hate crimes in the dataset are categorized according to a clear legal definition:

**Hate Crime:** A criminal offence committed against a person or property motivated in whole or in part by bias, prejudice, or hate based on race, national or ethnic origin, language, color, religion, sex, age, mental or physical disability, sexual orientation, or gender identity.

**Hate Incident:** A non-criminal action motivated by hate against an identifiable group, such as using racial slurs.

This data was obtained from Open Data Toronto through the **opendatatoronto** library (Gelfand 2022) \[\@gelfand2022\] All analysis was conducted in R (R Core Team 2023) \[\@r2023\] using packages such as **dplyr** (Wickham 2023) and **ggplot2** (Wickham 2016)\[\@wickham2016\]. The raw data features occurrences since 2018, providing valuable insight into hate crime trends in Toronto.

The dataset is intended to provide communities with information regarding public safety and awareness. However, it only includes confirmed hate crimes and not hate incidents.

## 2.2 Data Cleaning

Upon inspecting the dataset, it became clear that most of the data was already in good condition, as many variables were categorical in nature. To ensure consistency, we checked all the unique values within these categorical columns, such as `RACE_BIAS` and `RELIGION_BIAS`, and found no need for further standardization. Each category was consistently labeled, so no changes were necessary in terms of cleaning or formatting these values.

We also searched for missing values and duplicate rows. The dataset had very few missing values, and no duplicate entries were found. The only column that contained missing values was the `LOCATION_TYPE`, where some entries lacked information about where the hate crime occurred. To simplify the analysis, these missing values were replaced with the label "Unknown" instead of `NA`.

In addition, the date column `OCCURRENCE_DATE` and `REPORTED_DATE`—were originally in character format, so they were converted to proper date formats for convenience during analysis. Time columns, such as `OCCURRENCE_TIME` and `REPORTED_TIME`, were also reformatted into a standardized HH format.

Finally, we verified the `ARREST_MADE` column to ensure it contained only valid values, namely "YES" and "NO". Any invalid or ambiguous entries were replaced with `NA`.

These steps ensured that the dataset was clean and ready for further analysis, free of inconsistencies or issues that might have impacted the results.

## 2.3 Basic Summary Statistics of the Data

The summary table \@tbl-summary provides an overview of key variables from the hate crimes dataset, which spans from 2018 to 2023. The majority of incidents were reported in 2023. The most frequent occurrence locations were Streets/Roadways/Highways (287 incidents), with Division D32 handling the highest number of cases (177). A significant proportion of the crimes involved Race Bias, where Black individuals were the most frequently targeted (296 incidents), and Ethnicity Bias, with Chinese individuals being the most affected (40 incidents). Additionally, Religion Bias was notably prevalent against Jewish individuals, accounting for 485 out of 627 cases, while the most common Sexual Orientation Bias targeted 2SLGBTQ+ individuals (101 incidents).

In terms of primary offences, Mischief Under \$5000 was the most common offence, appearing in 605 incidents, while arrests were made in 270 cases, representing 20% of the total. Mischief Under \$5000 is a criminal offence involving willful damage or destruction of property valued at less than \$5000, such as vandalizing a car with spray paint.The dataset also tracks bias categories related to Age, Mental or Physical Disability, and Gender, with the majority of cases lacking a specific bias in these categories. This breakdown provides insight into the nature and distribution of hate crimes in Toronto over the five-year period.

| Column Name                   | Statistic                                                            |
|------------------------|-----------------------------------------------|
| REPORTED_YEAR                 | Most Frequent: 2023, Range: 2018–2023                                |
| LOCATION_TYPE                 | Most Frequent: Streets/Roadways/Highway, Count: 287                  |
| DIVISION                      | Most Frequent: D32, Count: 177                                       |
| OCCURRENCE_YEAR               | Most Frequent: 2023, Range: 2018–2023                                |
| AGE_BIAS                      | Total Count: 0                                                       |
| MENTAL_OR_PHYSICAL_DISABILITY | Total Count: 3                                                       |
| RACE_BIAS                     | Total Count: 428, Most Frequent: "Black", Count: 296                 |
| ETHNICITY_BIAS                | Total Count: 168, Most Frequent: "Chinese", Count: 40                |
| LANGUAGE_BIAS                 | Total Count: 3, Most Frequent: "Hindi", "Spanish", "Tamil", Count: 1 |
| RELIGION_BIAS                 | Total Count: 627, Most Frequent: "Jewish", Count: 485                |
| SEXUAL_ORIENTATION_BIAS       | Total Count: 197, Most Frequent: "2SLGBTQ+", Count: 101              |
| PRIMARY_OFFENCE               | Total Count: 605, Most Frequent: "Mischief Under \$5000", Count: 605 |
| GENDER_BIAS                   | Total Count: 78, Most Frequent: "Woman", Count: 29                   |
| ARREST_MADE                   | Total Arrest Made: 270                                               |

: Table: Summary of key variables {#tbl-summary}

### 2.4 Discussion of Data Selection

Most reports on hate crimes in Toronto, such as those cited in media outlets, often focus on general trends, like an increase in anti-Semitic hate crimes by a certain percentage, but they rarely provide details on the nature of the hate crimes themselves \[\@cbc2023\]. For example, they may report that anti-Semitic hate crimes rose by 56%, but they don’t specify whether these incidents involved threats, physical assaults, or vandalism.

To gain a more detailed understanding, this dataset was chosen to analyze not only the frequency of hate crimes but also the specific nature of the incidents. The goal was to answer questions like: What types of hate crimes are committed against different groups, such as Muslims or Jewish people? Are these crimes primarily verbal assaults, threats, or physical violence? This dataset allows for an in-depth analysis of the types of offences reported and whether law enforcement response, like arrests, varies by type of offence and targeted group.

## 3 Analysis

```{r}
library(tibble)


race_bias <- tibble(
  Group = c("Black", "East Asian", "South Asian", "White",  "Black, East Asian"),
  Count = c(296, 63, 22, 19, 7)
)



sexual_orientation_bias <- tibble(
  Group = c("2SLGBTQ+", "Gay", "Lesbian"),
  Count = c(101, 86, 9)
)



religion_bias <- tibble(
  Group = c("Jewish", "Muslim", "Catholic", "Sikh"),
  Count = c(485, 107, 12, 10)
)


ethnicity_bias <- tibble(
  Group = c("Chinese", "Indian", "Israeli", "Ukrainian", "Arab"),
  Count = c(40, 33, 17, 12, 18)
)


```

```{r}

plot_race_bias <- ggplot(race_bias, aes(x = reorder(Group, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Race Bias", x = "Group", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 9),
        axis.title.x = element_text(size = 8),  # Change font size for x-axis label
        axis.title.y = element_text(size = 8))  # Change font size for y-axis label




plot_sexual_orientation_bias <- ggplot(sexual_orientation_bias, aes(x = reorder(Group, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "burlywood1") +
  labs(title = "Sexual Orientation Bias", x = "Group", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 9),
        axis.title.x = element_text(size = 8),  # Change font size for x-axis label
        axis.title.y = element_text(size = 8))  # Change font size for y-axis label



plot_ethnicity_bias <- ggplot(ethnicity_bias, aes(x = reorder(Group, -Count), y = Count)) +
  geom_bar(stat = "identity",  fill = "lightgoldenrod1") +
  labs(title = "Ethnicity Bias", x = "Group", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 9),
        axis.title.x = element_text(size = 8),  # Change font size for x-axis label
        axis.title.y = element_text(size = 8))  # Change font size for y-axis label


plot_religion_bias <- ggplot(religion_bias, aes(x = reorder(Group, -Count), y = Count)) +
  geom_bar(stat = "identity", fill =  "cornsilk1") +
  labs(title = "Religion Bias", x = "Group", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 9),
        axis.title.x = element_text(size = 8),  # Change font size for x-axis label
        axis.title.y = element_text(size = 8))  # Change font size for y-axis label

grid.arrange(plot_race_bias, plot_ethnicity_bias, plot_religion_bias, plot_sexual_orientation_bias,
             ncol = 2,
             top = textGrob("Hate Crime Frequency by Victim Group and Bias Type", gp = gpar(fontsize = 11)))

```
